Parallel Sorting of Roughly-Sorted Sequences\\CSCI 5172 Fall '16 Project

### 2016-12-08 tex/paper.tex
my portion of the paper is finished

To determine the radius, we must first compute the $LR(A) = \{b_i\}$ and $RL(A) = \{c_i\}$ \textit{characteristic sequences}:
$$b_i = \max\{a_0, a_1, \cdots, a_i\}\,, \quad c_i = \min\{a_i, a_{i+1}, \cdots, a_{n-1}\}\,, \quad 0 \leq i < n$$
Both sequences are easily computed in linear time and space using simple min/max prefix scans.
Note that $c_i \leq b_i \;\forall\; i$.
By performing a linear-time scan of these sequences and observing where they ``cross'' each other, we can compute
  the \textit{disorder measure} sequence $DM(A) = \{d_i\}$, where each $d_i$ represents how displaced $a_i$ is from its position
  in the sorted ordering of the elements of $A$:
$$d_i = \max\big\{\{i - j \mid c_i < b_j\} \cup \{0\} \big\}$$

The radius of $A$ is thus the maximum element from $DM(A)$. We employ a modification of the $DM$ algorithm from
  \supercite{altman89} where we save memory by simply keeping track of the maximum-encountered $d_i$ instead of storing the
  entire sequence; our sequential radius determination algorithm thus runs in linear time and requires $2n = \Theta(n)$ space.

The core of the sequential Roughsort implementation is the aforementioned radius halving algorithm from \supercite{altman89}.
This algorithm halves the radius $k$ of $A$ in place through the following three steps:
\begin{enumerate}\setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}
\item Partition each consecutive run of $k$ elements in $A$ about the mean of the run.
\item Starting at element $a_{\lfloor k/2 \rfloor}$ to stagger the partitions, repeat step 1 and go to 3.
\item Repeat step 1 and halt.
\end{enumerate}
A run of elements is ``partitioned'' by partially sorting it such that no element before the median of the run exceeds
  the median and no element thereafter is less than it.
We partition each run using the \texttt{nth_element} function from the  C\texttt{++} STL, which performs an optimized,
  linear-time selection algorithm\cite{clrs} such as
  Introselect\footnote{Introselect is based on Quickselect, which resembles a Quicksort
  that searches for the $n$th element of the sorted array by only partitioning one side of each pivot.}\cite{selectalg}.

@@

We first attempted to implement the parallel approach described in \supercite{altman89}.
This approach follows the sequential version from Section 3, except in that the segment partitions are all launched in parallel
  at each of the 3 steps, synchronizing between steps.
We employed Nvidia's \textit{Thrust} library, which provides parallel CUDA implementations of the C\texttt{++} STL.

@@

Unfortunately, while this implementation was very simple to code and correctly sorted its input, it was thousands of times
  slower than the sequential implementation.
Running the implementation through CUDA's \texttt{nvvp} provider showed that the \texttt{sort} calls weren't being launched in
  parallel as expected, possibly due to some non-obvious synchronization mechanism that made concurrent launches of this
  function impossible.

We then implemented the parallel Roughsort approach described in \supercite{altman90}, which prescribes the parallel Mergesort
  algorithm by Cole in lieu of weak partitions and is implemented as follows for sorting the array $A$ with radius $k$ in place:
\begin{enumerate}\setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}
\item In parallel, sort all consecutive runs of $2k$ elements in $A$.
\item Repeat step 1, but begin the first run at $a_k$; halt.
\end{enumerate}

Unlike the sequential algorithm, this approach employs full sorting of the array segments and thus needs only two batches of
  sorts without intra-radial staggering.
Furthermore, just a single iteration of this process suffices to fully sort the $k$-sorted input.

We implemented this approach using a CUDA kernel where each thread in the launch is responsible for sorting exactly one segment
  from step 1 or 2; the steps are divided into separate kernel launches, each given an offset from which begin segmenting the
  array.
While the threads in each launch run concurrently\footnote{For large arrays, the GPU can't host every thread simultaneously and
  must sequentially execute the thread blocks in batches.}, each thread sorts its designated segment sequentially.
In contrast to \supercite{altman90}, we used Thrust's radix sort instead of its Mergesort implementation in order to leverage
  the linear runtime of the former.
While this approach to the parallel Roughsort was far more successful than the previous, it still competed more poorly than
  its competitors (see Section 9).

@@

\section{Implementation Performance Analysis}
mergesort speedup
upload times
ignore rad gen

@@

\section{Conclusion and Further Research}

>>>>

To determine the radius, we must first compute the \textit{characteristic sequences} $LR(A) = \{b_i\}$ and $RL(A) = \{c_i\}$:
$$b_i = \max\{a_0, a_1, \cdots, a_i\}\,, \quad c_i = \min\{a_i, a_{i+1}, \cdots, a_{n-1}\}\,, \quad 0 \leq i < n$$
Both sequences are easily computed in linear time and space using simple min/max prefix scans.
Note that $c_i \leq b_i \;\forall\; i$.
By performing a linear-time scan of these sequences and observing where they ``cross'' each other, we can compute
  the \textit{disorder measure} sequence $DM(A) = \{d_i\}$, where each $d_i$ represents how displaced $a_i$ is from its position
  in the sorted ordering of the elements of $A$:
$$d_i = \max\big\{\{i - j \mid c_i < b_j\} \cup \{0\} \big\}$$

The radius of $A$ is thus the maximum element from $DM(A)$. We employed a modification of the $DM$ algorithm from
  \supercite{altman89} where we save memory by simply keeping track of the maximum-encountered $d_i$ instead of storing the
  entire sequence; our sequential radius determination algorithm thus runs in linear time and requires $2n = \Theta(n)$ space.

The core of the sequential Roughsort implementation is the aforementioned radius halving algorithm from \supercite{altman89}.
This algorithm halves the radius $k$ of $A$ in place through the following three steps:
\begin{enumerate}\setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}
\item Partition each consecutive run of $k$ elements in $A$ about the mean of the run.
\item Starting at element $a_{\lfloor k/2 \rfloor}$ to stagger the partitions, repeat step 1 and go to 3.
\item Repeat step 1 and halt.
\end{enumerate}
A run of elements is ``partitioned'' by partially sorting it such that no element before the median of the run exceeds
  the median and no element thereafter is less than it.
We partition each run using the \texttt{nth_element} function from the  C\texttt{++} STL, which implements an optimized,
  linear-time selection algorithm\cite{clrs} such as
  Introselect\footnote{Introselect is based on Quickselect, which resembles a Quicksort
  that searches for the $n$th element of the sorted array by only partitioning one side of each pivot.}\cite{selectalg}.

@@

We first attempted to implement the parallel Roughsort described in \supercite{altman89}.
This approach follows the sequential version from Section 3, except in that the segment partitions are all launched in parallel
  at each of the 3 steps, synchronizing between steps.
We employed Nvidia's \textit{Thrust} library, which provides a parallel CUDA implementation of the C\texttt{++} STL.

@@

Unfortunately, while this implementation was very simple to code and correctly sorted its input, it was thousands of times
  slower than the sequential implementation.
Running the implementation through CUDA's \texttt{nvvp} profiler showed that the \texttt{sort} calls weren't being launched in
  parallel as expected, possibly due to some non-obvious synchronization mechanism that made concurrent launches of this
  function impossible.

We then implemented the parallel Roughsort approach described in \supercite{altman90}, which prescribes the parallel Mergesort
  algorithm by Cole in lieu of weak partitioning and is implemented as follows for sorting the array $A$ with radius $k$ in
  place:
\begin{enumerate}\setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}
\item In parallel, sort all consecutive runs of $2k$ elements in $A$.
\item Repeat step 1, but begin the first run at $a_k$; halt.
\end{enumerate}

Unlike the sequential algorithm, this approach fully sorts each array segment and thus needs only two batches of
  sorts without intra-radial staggering.
Furthermore, just a single iteration of this process suffices to fully sort the $k$-sorted input.

We implemented this approach using a CUDA kernel where each thread in the launch is responsible for sorting exactly one segment
  from step 1 or 2; the steps are divided into separate kernel launches, each given an offset from which begin segmenting the
  array.
While the threads in each launch run concurrently\footnote{For large arrays, the GPU can't host every thread simultaneously and
  must sequentially execute the thread blocks in batches.}, each thread sorts its designated segment sequentially.
In contrast to \supercite{altman90}, we used Thrust's radix sort instead of its Mergesort implementation in order to leverage
  the linear runtime of the former.
While this approach to the parallel Roughsort was far more successful than the previous, it still struggled against
  its competitors (see Section 9).

@@

\section{Implementation Performance Analysis}
We tested our sequential and parallel Roughsort implementations against sequential and parallel Mergesort (as implemented by
  the C\texttt{++} STL and by CUDA Thrust, respectively) as well as against our simple, sequential Bubblesort implementation.
We chose Mergesort since it's a classic sort algorithm that also scales very well with parallel execution, capable
  of achieving a speedup of at least $\Theta(n / \lg^2 n)$ when given unlimited processors for a runtime of
  $\Theta(\lg^3 n)$\cite{clrs}.
While Bubblesort has a quadratic runtime upper bound, its lightweight, linear lower bound emerges when given nearly-sorted
  sequences and provides a means of comparing our implementation's performance for small values of $k$.

All tests were run on a workstation sporting a high-end, quad-core Intel CPU as well as an Nvidia GeForce GTX 1080---a
  recent and high-end\footnote{At the time of this writing.}, consumer-grade GPU suitable for heavy computational workloads
  over 32-bit values.

The time needed to transfer data to the GPU prior to running the parallel sorts isn't counted in the runtime results, since we'd
  expect such sorting operations to run as part of larger GPU workloads and not in isolation.
Since our parallel radius determiner wasn't complete until late in development and ran slower than the sequential
  implementation over arrays with larger radii, our parallel Roughsort implementation used the sequential implementation to
  compute $k$, the runtime of which wasn't figured into the results.

Figures 3 and 4 compare all five sorting algorithms, varying $n$ while fixing $k$ to 2 and 15, respectively. Roughsort ought
  to have a runtime advantage with the smaller radius that dissipates as we increase $k$.
Figures 5 and 6 invert the axes and vary $k$ while fixing $n$ at $750 \cdot 10^3$ and $1.25 \cdot 10^6$, respectively. 
For each tested $(n, k)$, all five sorters were run over the same, $k$-sorted random array of length $n$; their runtimes were
  averaged over eight such runs.
The runtime axis is logarithmic in every graph.

Figure 3 demonstrates the utility of the sequential Roughsort algorithm, which beats even the highly-optimized Mergesort from
  the C\texttt{++} STL at handling the nearly-sorted input arrays.
This victory is greatly overshadowed by the sequential Bubblesort, which is operating close to its optimal domain.

The next battery of tests in Figure 4 raises $k$ to 15, causing the sequential Roughsort's advantage over Mergesort to evaporate.
The parallel Roughsort generally outperforms Bubblesort here (which fairs better than expected against the sequential Mergesort
  in a trend that can't last for much longer to the right of the graph) and defeats the sequential Roughsort.
Unexpectedly, the parallel implementation exhibits about a factor-of-ten speedup over the $k = 2$ graph; this shows that
  the overhead of launching many CUDA threads over small array subsequences substantially counterbalances the greater runtime
  of sorting larger subsequences.

Figures 5 and 6 fix $n$ and vary $k$, yielding similar results.
For arrays of these lengths, the parallel Roughsort's crossover point lies somewhere around the interval $20 \leq k \leq 30$,
  where it manages to best the three sequential algorithms.
An invariant among all four graphs is the utter dominance of the parallel Mergesort implementation provided by Thrust,
  whose runtimes never depart from the $x$-axis.

These results suggest that our parallel Roughsort's constant runtime factor and crossover point are relatively high.
We were unable to observe a benefit from its superior asymptotic bound over nearly-sorted arrays that could fit within the
GPU's memory.

@@

\section{Conclusion and Further Research}
An implementation of Cole's parallel median algorithm might enable a more efficient parallel Roughsort\cite{cole85}.
